---
title: "Module 1: Climate change"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}
#| warning: false
library(tidyverse)
theme_set(theme_classic())
```

### Use of GitHub

Link to your forked GH repository: https://github.com/tbaugh323/Module1-climate

### Use of Quarto

Link to your .qmd file: https://github.com/tbaugh323/Module1-climate/blob/main/module1.qmd 

## CO2 trends from Mauna Loa

We will begin at the global scale by plotting the famed Mauna Loa CO~2~ curve at monthly intervals, with the twin objectives of (1) making layered plots with 'ggplot2', a package within the 'tidyverse' collection widely adopted by the R data science community, and (2) refreshing understanding of the seasonal cycle in atmospheric \[CO~2~\].

With data you've collected yourself, it is more common to store locally as a csv. However, we can use tidyverse tools to read directly from files stored online and will do so here to get the latest data.

```{r}
#| warning: false
#| output: false
co2 <-read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date", "average",
                                "interpolated", "trend", "days"),
                  na = c("-1", "-99.99"))
```

You can examine these data in several ways. One option is to use your mouse in the Environment tab and click on the dataset name. Alternatively, you can type `View(co2)` in your console to achieve the same outcome. Below are some other common options to understand the data structure at a glance.

```{r}
str(co2)
head(co2)
```

### Q1 (1 point)

Create a timeseries plot of average \[CO~2~\]. Use `as.Date()` to create a formatted date for the x-axis. Label both the y-axis and add a line to connect the points.

```{r}
co2 <- co2 |> 
  mutate(year = as.character(year), month = as.character(month),
         date = as.Date(paste0(year, "-" , month, "-15"), format = "%Y-%m-%d"))

co2 |> 
  ggplot(aes(x = date, y = average)) +
  geom_point() +
  geom_line(color = "red") +
  labs(y = "Monthly Average CO2 (ppmv)", x = "Date")
```

### Q2 (1 point)

In which 3-month window are the \[CO~2~\] values at their seasonal peak? Create another dataframe with these months selected and create a plot like above, but with the maximum months colored in "forestgreen".

```{r}
co2_winter <- co2 |> 
  filter(month %in% c(4, 5, 6))

co2 |> 
  ggplot(aes(x = date, y = average)) +
  geom_point() +
  geom_line(color = "red") +
  geom_point(data = co2_winter, aes(x = date, y = average), color = "forestgreen") +
  labs(y = "Monthly Average CO2 (ppmv)", x = "Date")
```

### Q3 (1 point)

Why does \[CO~2~\] peak in the months highlighted above? Consider the location of Mauna Loa and the seasonality of processes that emit and absorb \[CO~2~\].  

After reading Lewandowsky et al. 2016, how might you go about separating the long term trend and the seasonal variation in \[CO~2~\]?

\[CO~2~\] peaks in April, May, and June because this is right before the start of the growing season for most plants in the Northern Hemisphere. Therefore, the plants have been "off" for the maximum amount of time right before the summer, so \[CO~2~\] has been increasing for the maximum amount of time. Like Lewandowsky et al., you could do Z-scores to help quantify deviation. You could also do a yearly average instead of monthly averages, with error bars indicating variation from seasonality. This would help highlight multi year trends rather than intra year variation. 

## Global temperature

Next, we turn to a global temperature anomaly [dataset](https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt) from NASA GISS, similar to the GMST used by Lewandowsky et al. 2016.

### Q4 (1 point)

Read some [background](https://data.giss.nasa.gov/gistemp/history/background.html) as well as the [simplified documentation](https://climate.nasa.gov/vital-signs/global-temperature/?intent=121) for this dataset.

How are the measurements made? Describe each column in the dataset and its units. What is the temporal resolution of the data?

They sampling air temperature at several weather stations in geographically diverse locations around the world. Year is the year of measurements. There are two temperature columns describing temperature anomaly. The Lowess smoothing column are data which have been smoothed to remove annual variability, where the No_Smoothing column has no smoothing. Both temperature columns are in Celsius, and their temporal resolution is yearly.

### Q5 (1 point)

Construct code to import the dataset from the above url in a format that can be used for transforming and visualization.

```{r}
temp <-read_table("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt", 
                  skip = 5,
                  col_names = c("Year", "No_Smoothing", "Lowess"))
```

### Q6 (1 point)

Plot the trend in global mean temperature over time. Describe what you see and how you interpret the patterns you observe.

```{r}
temp |> 
  ggplot(aes(x = Year, y = No_Smoothing)) +
  geom_point() +
  labs(x = 'Date', y = 'Temperature Anomaly (C)')
```

The annual temperature anomaly begins to increase just after 1960. There is a decent amount of variability year-to-year but there is still a very clear upward trend in the more recent years. 

### Q7 (1 point)

Consider the "pause" analyzed in Lewandowsky et al. 2016. Do you see evidence for such a pause?

We do not see evidence for the supposed pause considered in Lewandowsky et al. 2016. Annual temperature anomaly is clearly and consistently increasing without a pause for the past 50 years. The variability year-to-year is not significant when compared to the long-term trend in the dataset.

### Bonus 1 (1 point)

Recreate the analysis from Lewandowsky's Fig. 1.

## Long term records of CO2

The data from Mauna Loa go back as far as modern instrumentation. To put these values into geological perspective, we must go back much further than humans have been monitoring \[CO~2~\]. Ice sheets contain layers of snow with trapped air bubbles and can be cored and dated like tree rings.

Previously hosted by Oak Ridge National Lab, the [data](https://zenodo.org/records/3678928) associated with the famed Vostok Ice Core is now archived on Zenodo. Skim [Barnola et al. 1987](https://www.nature.com/articles/329408a0) as needed.

### Q8 (1 point)

After downloading this dataset, create a folder entitled `data` and save the file within the folder with a .txt extension. Read this file into R. What are the columns and units? In the "Vostok CO2 record" section of Barnola et al. 1987, what is the temporal resolution of the data?

```{r}
co2_record <- read_delim('data/vostok.icecore.txt',
                         skip = 21,
                         col_names = c('depth', 'ice_age', 'air_age', 'co2'),
                         delim = '\t') |> 
  mutate(co2 = str_split_i(air_age, '\t', 2),
         co2 = ifelse(is.na(co2), str_split_i(air_age, '   ', 2), co2),
         co2 = ifelse(is.na(co2), str_split_i(air_age, '  ', 2), co2),
         air_age = str_split_i(air_age, '\t', 1),
         air_age = ifelse(grepl('   ', air_age), str_split_i(air_age, '   ', 1), air_age),
         air_age = ifelse(grepl('  ', air_age), str_split_i(air_age, '  ', 1), air_age))

```

There is a depth column with units of meters, an age of ice column with units of years BP, an age of air column with units of years BP, and a co2 concentration column with units of ppmv. The temporal resolution is around 300-750 years.

### Q9 (1 point)

Transform the data as needed to create a plot in chronological order. Add at least two different smoothing window averages to your plot. How did you select the smoothing windows?

```{r}
bin_data <- function(bin_size) {
  year_bins <- seq(0, 450000, bin_size)
  mean_co2 <- c()
  
  for(i in (1:(length(year_bins) - 1))){
    hold <- co2_record |> 
      filter(air_age > year_bins[i] & air_age <= year_bins[i+1])
  
    mean_co2 <- append(mean_co2, mean(as.numeric(hold$co2)))
  }
  
  binned_co2 <- data.frame(year_bins[-1] - bin_size / 2, mean_co2) |> 
    filter(!is.na(mean_co2))
  
  colnames(binned_co2) <- c('year_bins', 'mean_co2')
  
  binned_co2
}

co2_5ka_record <- bin_data(5000)
co2_10ka_record <- bin_data(10000)

co2_record |> 
  mutate(co2 = as.numeric(co2),
         air_age = as.numeric(air_age)) |> 
  ggplot(aes(x = air_age, y = co2)) + 
  geom_point() + 
  scale_x_reverse() +
  geom_point(data = co2_5ka_record, aes(x = year_bins, y = mean_co2), color = 'tomato') +
  geom_line(data = co2_5ka_record, aes(x = year_bins, y = mean_co2), color = 'tomato') +
  geom_point(data = co2_10ka_record, aes(x = year_bins, y = mean_co2), color = 'blue') +
  geom_line(data = co2_10ka_record, aes(x = year_bins, y = mean_co2), color = 'blue') +
  annotate('text', x = 85000, y = 280, label = '5,000 year average', color = 'tomato') + 
  annotate('text', x = 85000, y = 290, label = '10,000 year average', color = 'blue') + 
  annotate('text', x = 85000, y = 300, label = 'original data', color = 'black') + 
  labs(x = 'Age of Air (years)', y = 'CO2 concentation (ppmv)')

```

We chose our smoothing window by trying out a range of values and seeing what sizes over or undersmoothed the data. We settled on two values that we thought provided sufficient smoothing while still capturing the general trends present in the data.

### Q10 (1 point)

Join your data with the Mauna Loa data and plot together, considering different colors for the different datasets.

```{r}
co2 <- co2 |> 
  mutate(timestamp = 1987 - as.numeric(decimal_date))

t <- c(co2_record$air_age, co2$timestamp)
c <- c(co2_record$co2, co2$average)
ds <- c(rep('ice_core', nrow(co2_record)), rep('measurements', nrow(co2)))

joined_data <- data.frame(t, c, ds) |> 
  mutate(t = as.numeric(t),
         c = as.numeric(c))

joined_data |> 
  ggplot(aes(x = t, y = c, color = ds)) + 
  geom_point() +
  scale_x_reverse() + 
  labs(x = 'years before 1987', 
       y = 'CO2 concentration (ppmv)')

joined_data <- joined_data |> 
  mutate(ds = case_when(ds == "ice_core" ~ "Historic data",
                        ds == "measurements" ~ "Anthropogenic time"))

joined_data |> 
  ggplot(aes(x = t, y = c, color = ds)) + 
  geom_point() +
  scale_x_reverse() + 
  labs(x = 'Years before 1987', 
       y = expression(paste(CO[2], " Concentration (ppmv)")),
       color = "Dataset") +
  scale_color_manual(values = c("tomato", "plum")) +
  theme(axis.text = element_text(size = 13),
        axis.title = element_text(size = 15),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 15))
```


### Bonus 2 (1 point)

Write a catchy title and short description in the style of a NYtimes infographic, contrasting the rate of anthropogenic vs. background climate change for a general public audience.

Title: Cataclysmic Carbon Climb
Description: Carbon emissions (measured in parts per million by volume) from two different datasets. Historic data (prior to ~2000 years ago) are derived from ice core records, in which carbon concentration in the air is captured in frozen water. Recent measurements (including data from the twentieth and twenty-first centuries) come from the Mauna Loa Observatory, and are measured with modern instruments. Recent (Mauna Loa) data are inferred to be affected by anthropogenic carbon emissions, whereas long term historic climate trends are captured in ice core data. Carbon emissions in recent decades spikes considerably compared to the historic averages.
